---
title: "MA 679 Final Project"
author: "Chunlin Liu"
date: "2026-04-25"
output:
  pdf_document: default
  html_document: default
  word_document: default
---


EDA


```{r setup, include=FALSE}
library(foreign)
library(haven)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tableone)
library(MASS)
library(car)


Sleeping_length_pre <- read_xpt("P_SLQ.xpt.txt")
Sleeping_length_post <- read_xpt("SLQ_L.xpt.txt")

Demographic_pre <- read_xpt("P_DEMO.xpt.txt")
Demographic_post <- read_xpt("DEMO_L.xpt.txt")

Depression_pre <- read_xpt("P_DPQ.xpt.txt")
Depression_post <- read_xpt("DPQ_L.xpt.txt")


Sleep_all <- bind_rows(Sleeping_length_pre, Sleeping_length_post)
Demographic_all  <- bind_rows(Demographic_pre, Demographic_post)
Depress_all <- bind_rows(Depression_pre, Depression_post)

nhanes_all <- Depress_all %>%
  left_join(Sleep_all, by = "SEQN") %>%
  left_join(Demographic_all, by = "SEQN")



# function to convert NHANES missing codes
clean_nhanes_missing <- function(x) {
  x <- ifelse(x %in% c(77, 99, 777, 9999, 77777, 99999), NA, x)
  return(x) 
}


colSums(is.na(nhanes_all))


vars_to_keep <- c(
  "SEQN",
  "SLQ300", "SLQ310", "SLQ320", "SLQ330",
  "SLD012", "SLD013",
  paste0("DPQ0", c("10","20","30","40","50","60","70","80","90")),
  "DPQ100",
  "RIDAGEYR", "RIAGENDR", "RIDRETH3", "DMDEDUC2","DMDMARTZ","RIDEXPRG", "INDFMPIR","DMDBORN4", "DMDMARTZ", "INDFMPIR"
)



data_raw <- nhanes_all %>%
  dplyr::select(all_of(vars_to_keep))


data_raw <- data_raw %>%
  mutate(
    pregnant = case_when(
      RIDEXPRG == 1 ~ "Pregnant",
      TRUE ~ "NotPregnant"
    )
  )

data_raw <- data_raw %>%
  mutate(
    marital3group = case_when(
      DMDMARTZ == 1 ~ "Partnered",
      DMDMARTZ == 2 ~ "PreviouslyMarried",
      DMDMARTZ == 3 ~ "NeverMarried",
      TRUE ~ "Unknown"
    )
  )


colSums(is.na(data_raw))

#re structure data on sleeping category
df <- data_raw %>%
  mutate(
    sleep_hours = SLD012,
    sleep_cat = case_when(
      sleep_hours < 6 ~ "<6h",
      sleep_hours >= 6 & sleep_hours <= 8 ~ "6-8h",
      sleep_hours > 8 ~ ">8h",
      TRUE ~ NA_character_
    )
  )


# also do the re structure of education level

df <- df %>%
  mutate(
    edu_level = case_when(
      DMDEDUC2 == 1 ~ "Less than 9th",
      DMDEDUC2 == 2 ~ "9-11th (no diploma)",
      DMDEDUC2 == 3 ~ "High school/GED",
      DMDEDUC2 == 4 ~ "Some college/AA",
      DMDEDUC2 == 5 ~ "College or above",
      TRUE ~ "Unknown"
    ),
    edu_level = factor(edu_level,
                       levels = c("Less than 9th",
                                  "9-11th (no diploma)",
                                  "High school/GED",
                                  "Some college/AA",
                                  "College or above",
                                  "Unknown"))
  )



table(df$edu_level)


ggplot(df, aes(x = edu_level)) +
  geom_bar() +
  labs(
    title = "Education Level Distribution",
    x = "Education Level",
    y = "Count"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# use the sum of depression to categorize grades
df <- df %>%
  mutate(
    phq9 = DPQ010 + DPQ020 + DPQ030 + DPQ040 + DPQ050 +
           DPQ060 + DPQ070 + DPQ080 + DPQ090,
    depressed_score = ifelse(phq9 >= 10, 1, 0)
  )

df$sleep_cat <- factor(df$sleep_cat,
                       levels = c("<6h", "6-8h", ">8h"))



# variables to check
vars <- c("DMDBORN4", "DMDMARTZ", "RIDEXPRG", "INDFMPIR")

# calculate missing counts and missing rates
missing_summary <- data.frame(
  Variable = vars,
  Missing_Count = sapply(Demographic_all[vars], function(x) sum(is.na(x))),
  Total_N = nrow(Demographic_all),
  Missing_Rate = sapply(Demographic_all[vars], function(x) mean(is.na(x)))
)

# convert to percentage
missing_summary$Missing_Rate_Percent <- round(
  missing_summary$Missing_Rate * 100, 2
)

missing_summary



# keep only people eligible for marital status question
adult20 <- Demographic_all %>%
  filter(RIDAGEYR >= 20)

# recode marital status
adult20 <- adult20 %>%
  mutate(
    marital_status = case_when(
      DMDMARTZ == 1 ~ "Married/Living with partner",
      DMDMARTZ == 2 ~ "Widowed/Divorced/Separated",
      DMDMARTZ == 3 ~ "Never married",
      is.na(DMDMARTZ) ~ "Missing"
    )
  )

# summary
marital_summary <- adult20 %>%
  count(marital_status) %>%
  mutate(
    percent = round(n / sum(n) * 100, 2)
  )

# missing rate among eligible people
missing_rate <- mean(is.na(adult20$DMDMARTZ))

cat("Missing rate among age >= 20:",
    round(missing_rate * 100, 2), "%\n\n")

marital_summary

```

EDA

```{r}
CreateTableOne(vars = c("sleep_hours","phq9","RIDAGEYR","RIAGENDR","DMDEDUC2","DMDBORN4", "DMDMARTZ", "INDFMPIR"),
               strata = "sleep_cat", data = df, test = FALSE)


# sleep hours
ggplot(df, aes(sleep_hours)) +
  geom_histogram(bins = 24) +
  labs(title = "Distribution of Sleep Hours")



#depression

  #shown in grades
ggplot(df, aes(phq9)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of PHQ-9 Score")

  # log phq9
df <- df %>%
  mutate(
    log_phq9 = log(phq9 )
  )

ggplot(df, aes(x = log_phq9)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Log-Transformed PHQ-9 Score",
    x = "log(PHQ-9)",
    y = "Count"
  ) +
  theme_minimal()


  #shown in groups
df <- df %>%
  mutate(
    phq_cat = case_when(
      phq9 < 5 ~ "None",
      phq9 < 10 ~ "Mild",
      phq9 < 15 ~ "Moderate",
      TRUE ~ "Severe"
    ),
    phq_cat = factor(phq_cat, ordered = TRUE)
  )


df %>%
  count(phq_cat) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ggplot(aes(x = phq_cat, y = percent)) +
  geom_col() +
  labs(
    title = "Distribution of PHQ-9 Categories",
    x = "Depression Category",
    y = "Percent (%)"
  )




# when to go to bed

library(lubridate)
df <- df %>%
  mutate(
    bed_time_weekday = hm(SLQ300),
    wake_weekday = hm(SLQ310),
    bed_time_weekend = hm(SLQ320),
    wake_weekend = hm(SLQ330)
  )


df <- df %>%
  mutate(
    bed_time_weekday_hour = hour(bed_time_weekday) + minute(bed_time_weekday)/60,
    bed_time_weekend_hour = hour(bed_time_weekend) + minute(bed_time_weekend)/60
  )



# bed time on weekdays
ggplot(df, aes(x = bed_time_weekday_hour)) +
  geom_histogram(bins = 24) +
  labs(
    title = "Weekday Bedtime",
    x = "Hours",
    y = "Count"
  )

#bed time on weekends


ggplot(df, aes(x = bed_time_weekend_hour)) +
  geom_histogram(bins = 24) +
  labs(
    title = "Weekend Bedtime",
    x = "Hours",
    y = "Count"
  )

# depression grades different by sleeping categorys
ggplot(df, aes(x = sleep_cat, y = phq9)) +
  geom_boxplot()

#table for depression and sleep length
df %>%
  group_by(sleep_cat) %>%
  summarise(
    depression_score = mean(depressed_score, na.rm = TRUE)
  )


ggplot(df, aes(x = phq_cat, fill = sleep_cat)) +
  geom_bar(position = "fill") +
  labs(
    title = "PHQ-9 Category by Sleep Length",
    x = "Depression Category",
    y = "Proportion"
  )

# heatmap for depression and sleeping length

library(dplyr)
library(ggplot2)
library(reshape2)

# Select variables for heatmap
df_heat <- df %>%
  dplyr::select(
    depressed_score,
    sleep_cat,
    bed_time_weekday_hour,
    bed_time_weekend_hour,
    RIDAGEYR,
    RIAGENDR,
    RIDRETH3,
    DMDEDUC2,
    DMDMARTZ,
    INDFMPIR,
    DMDBORN4
  ) %>%
  mutate(across(everything(), ~ as.numeric(as.factor(.))))

# Correlation matrix
cor_matrix <- cor(
  df_heat,
  method = "spearman",
  use = "pairwise.complete.obs"
)

# Heatmap data

library(reshape2)
cor_df <- melt(cor_matrix)

# Plot heatmap
ggplot(cor_df, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  labs(
    title = "Spearman Correlation Heatmap",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )


#table for depression and gender
df %>%
  group_by(RIAGENDR) %>%
  summarise(
    mean_sleep = mean(sleep_hours, na.rm=TRUE),
    mean_phq9 = mean(phq9, na.rm=TRUE)
  )

#summary for depression(binomial) and depression phq9(sum of score)
summary(df$depressed_score)
summary(df$phq9)


#check missing data
library(dplyr)
missing_table <- df %>%
  summarise(across(
    everything(),
    ~ sum(is.na(.))
  )) %>%

  tidyr::pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Missing_Count"
  ) %>%
  mutate(
    Total_N = nrow(df),
    Missing_Rate = round(Missing_Count / Total_N * 100, 2)
  ) %>%
  arrange(desc(Missing_Rate))
missing_table


```


modeling











# redge and lasso for dimention reduction


```{r}


library(dplyr)
library(tidyr)
library(glmnet)

# redge
nhanes_ml <- nhanes_all %>%
  filter(RIDAGEYR >= 20) %>%
  mutate(
    # create PHQ-9 total score
    phq9 = DPQ010 + DPQ020 + DPQ030 +
      DPQ040 + DPQ050 + DPQ060 +
      DPQ070 + DPQ080 + DPQ090,
    
    # create sleep hours from raw NHANES variable
    sleep_hours = SLD012,
    
    # binary outcome
    depressed = ifelse(phq9 >= 10, 1, 0),
    
    # sleep category
    sleep_cat = case_when(
      sleep_hours < 6 ~ "<6h",
      sleep_hours <= 8 ~ "6-8h",
      sleep_hours > 8 ~ ">8h"
    ),
    
    # marital status
    marital3group = case_when(
      DMDMARTZ == 1 ~ "Married/Partner",
      DMDMARTZ == 2 ~ "Previously Married",
      DMDMARTZ == 3 ~ "Never Married"
    )
  ) %>%
  dplyr::select(
    depressed,
    sleep_cat,
    RIDAGEYR,
    RIAGENDR,
    RIDRETH3,
    DMDEDUC2,
    marital3group,
    INDFMPIR,
    DMDBORN4
  ) %>%
  drop_na()

# predictor matrix

x <- model.matrix(
  depressed ~ .,
  data = nhanes_ml
)[, -1]

# outcome
y <- nhanes_ml$depressed


# Ridge logistic regression

set.seed(123)
#ridge_cv <- cv.glmnet(
#  x,
#  y,
#  alpha = 0,     
#  family = "binomial"

#)

# plot CV error

#plot(ridge_cv)

# best lambda

#ridge_cv$lambda.min
#ridge_cv$lambda.1se
# coefficients at lambda.min
#coef(ridge_cv, s = "lambda.min")

# coefficients at lambda.1se
#coef(ridge_cv, s = "lambda.1se")





# lasso


 create predictor matrix
x <- model.matrix(
  depressed ~ .,
  data = nhanes_ml
)[, -1]

 outcome
y <- nhanes_ml$depressed

 fit cross-validated LASSO logistic regression
set.seed(123)

lasso_cv <- cv.glmnet(
  x,
  y,
  alpha = 1,
  family = "binomial"
)

# plot CV error
plot(lasso_cv)

# coefficients at lambda.min
coef(lasso_cv, s = "lambda.min")

# coefficients at lambda.1se
coef(lasso_cv, s = "lambda.1se")

```


In the final model, I would keep variables: sleep_cat, RIDAGEYR, RIAGENDR, DMDEDUC2, marital3group, INDFMPIR, and DMDBORN4. 


```{r}
library(dplyr)
library(lubridate)

nhanes_ml <- nhanes_all %>%
  filter(RIDAGEYR >= 20) %>%
  
  mutate(
    
    phq9 = DPQ010 + DPQ020 + DPQ030 +
      DPQ040 + DPQ050 + DPQ060 +
      DPQ070 + DPQ080 + DPQ090,
    phq_cat = case_when(
      phq9 < 5 ~ "None",
      phq9 < 10 ~ "Mild",
      phq9 < 15 ~ "Moderate",
      phq9 >= 15 ~ "Severe",
      TRUE ~ NA_character_
    ),
    
    depressed = ifelse(phq9 >= 10, 1, 0),
    
    sleep_hours = SLD012,
    
    sleep_cat = case_when(
      sleep_hours < 6 ~ "<6h",
      sleep_hours <= 8 ~ "6-8h",
      sleep_hours > 8 ~ ">8h"
    ),
    
    # convert character times like "22:30"
    bed_time_weekday =
      hm(as.character(SLQ300)),
    
    bed_time_weekend =
      hm(as.character(SLQ320)),
    
    bed_time_weekday_hour =
      hour(bed_time_weekday) +
      minute(bed_time_weekday) / 60,
    
    bed_time_weekend_hour =
      hour(bed_time_weekend) +
      minute(bed_time_weekend) / 60,
    
    marital3group = case_when(
      DMDMARTZ == 1 ~ "Married/Partner",
      DMDMARTZ == 2 ~ "Previously Married",
      DMDMARTZ == 3 ~ "Never Married"
    )
    
  ) %>%
  
  dplyr::select(
    depressed,
    phq9,
    phq_cat,
    sleep_cat,
    bed_time_weekday_hour,
    bed_time_weekend_hour,
    RIDAGEYR,
    RIAGENDR,
    DMDEDUC2,
    marital3group,
    INDFMPIR,
    DMDBORN4
  ) %>%
  
  drop_na()


```









# split training and testing data

```{r}
library(caret)


set.seed(123)

# create train/test split
train_index <- createDataPartition(
  nhanes_ml$depressed,
  p = 0.7,
  list = FALSE
)

# training data
train_data <- nhanes_ml[train_index, ]

# testing data
test_data <- nhanes_ml[-train_index, ]

# check sizes
dim(train_data)
dim(test_data)

# check outcome balance
prop.table(table(train_data$depressed))
prop.table(table(test_data$depressed))
```

# corss validation

```{r}
library(caret)

# make sure outcome is factor
train_data$depressed <- factor(
  train_data$depressed,
  levels = c(0, 1)
)

# 5-fold cross validation setup
ctrl <- trainControl(
  method = "cv",
  number = 5
)

# logistic regression with CV
cv_log_model <- train(
  depressed ~
    sleep_cat +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,
  
  data = train_data,
  method = "glm",
  family = "binomial",
  trControl = ctrl
)

# results
cv_log_model
```







# binary outcome model

```{r}
library(dplyr)

library(caret)

# binary logistic regression model

log_model <- glm(
  depressed ~
    sleep_cat +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,

  family = binomial,

  data = train_data

)





summary(log_model)

# predicted probabilities



pred_prob <- predict(

  log_model,

  newdata = test_data,

  type = "response"

)

# convert probabilities to classes

pred_class <- ifelse(pred_prob > 0.5, 1, 0)

# confusion matrix

confusionMatrix(

  factor(pred_class),

  factor(test_data$depressed)

)
```








Comparing these two model, incuding the preduction accuracy and AIC, binary model is a better choice. 


# check ROC-AUC plot

```{r}
# predicted probabilities from logistic model
pred_prob <- predict(
  log_model,
  newdata = test_data,
  type = "response"
)

# create prediction classes at many thresholds
thresholds <- seq(0, 1, by = 0.01)

roc_data <- data.frame()

for (t in thresholds) {
  
  pred_class <- ifelse(pred_prob >= t, 1, 0)
  
  TP <- sum(pred_class == 1 & test_data$depressed == 1)
  TN <- sum(pred_class == 0 & test_data$depressed == 0)
  FP <- sum(pred_class == 1 & test_data$depressed == 0)
  FN <- sum(pred_class == 0 & test_data$depressed == 1)
  
  sensitivity <- TP / (TP + FN)
  specificity <- TN / (TN + FP)
  
  roc_data <- rbind(
    roc_data,
    data.frame(
      threshold = t,
      sensitivity = sensitivity,
      specificity = specificity
    )
  )
}

# false positive rate
roc_data$fpr <- 1 - roc_data$specificity


auc <- sum(
  diff(roc_data$fpr) *
    (head(roc_data$sensitivity, -1) +
       tail(roc_data$sensitivity, -1)) / 2
)

# ROC curve
ggplot(roc_data, aes(x = fpr, y = sensitivity)) +
  geom_line(linewidth = 1) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 0.7,
    y = 0.2,
    label = paste("AUC =", round(auc, 3)),
    size = 5
  ) +
  labs(
    title = "ROC Curve for Logistic Regression",
    x = "False Positive Rate",
    y = "True Positive Rate"
  ) +

  theme_minimal()



```


# decision tree
```{r}
library(tree)

# fit classification tree
tree_model <- tree(
  factor(depressed) ~
    sleep_cat +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,
  
  data = train_data
)

# summary
summary(tree_model)

# plot tree
plot(tree_model)
text(tree_model, pretty = 0)

# predictions
tree_pred <- predict(
  tree_model,
  newdata = test_data,
  type = "class"
)

# confusion matrix
table(
  Predicted = tree_pred,
  Actual = test_data$depressed
)


# use bed time to do tree



train_balanced <- train_data %>%
  group_by(depressed) %>%
  sample_n(min(table(train_data$depressed))) %>%
  ungroup()


tree_model_bal <- tree(
  factor(depressed) ~
    sleep_cat +
    bed_time_weekday_hour +
    bed_time_weekend_hour +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,
  
  data = train_balanced
)

plot(tree_model_bal)
text(tree_model_bal, pretty = 0)


```

# random forest

```{r}
library(randomForest)

set.seed(123)

# random forest model
rf_model <- randomForest(
  factor(depressed) ~
    sleep_cat +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,
  
  data = train_data,
  
  ntree = 500,
  importance = TRUE
)

# model summary
print(rf_model)

# variable importance
importance(rf_model)

# importance plot
varImpPlot(rf_model)

# predictions
rf_pred <- predict(
  rf_model,
  newdata = test_data
)

# confusion matrix
table(
  Predicted = rf_pred,
  Actual = test_data$depressed
)





# Create balanced training data


table(train_data$depressed)

set.seed(123)

train_balanced <- train_data %>%
  group_by(depressed) %>%
  sample_n(min(table(train_data$depressed))) %>%
  ungroup()

# check balance
table(train_balanced$depressed)


# Refit Random Forest


set.seed(123)

rf_model_bal <- randomForest(
  factor(depressed) ~
    sleep_cat +
    bed_time_weekday_hour +
    bed_time_weekend_hour +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,
  
  data = train_balanced,
  
  ntree = 500,
  importance = TRUE
)

# model summary
print(rf_model_bal)


# Variable Importance


importance(rf_model_bal)

varImpPlot(rf_model_bal)


# Predictions on ORIGINAL test set


rf_pred_bal <- predict(
  rf_model_bal,
  newdata = test_data
)

# confusion matrix
confusionMatrix(
  factor(rf_pred_bal),
  factor(test_data$depressed)
)

```

Sleep duration category alone may have limited predictive contribution compared with socioeconomic and demographic variables. Accuracy can be misleading under class imbalance

Refitting the model, the accuracy decreased. However, it is a bias variance trade off. Although the overall accuracy became lower, the balanced model improved its ability to identify depressed individuals rather than simply predicting the majority class. As a result, sensitivity and balanced accuracy improved, leading to a more meaningful and clinically useful classification model.



# SVM
```{r}
library(e1071)
library(caret)

# make sure factor outcome
train_balanced$depressed <- factor(
  train_balanced$depressed,
  levels = c(0, 1)
)

test_data$depressed <- factor(
  test_data$depressed,
  levels = c(0, 1)
)



# Linear SVM


set.seed(123)

svm_linear <- svm(
  depressed ~
    sleep_cat +
    bed_time_weekday_hour +
    bed_time_weekend_hour +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,

  data = train_balanced,
  kernel = "linear",
  cost = 1,
  scale = TRUE

)



# predictions

svm_linear_pred <- predict(
  svm_linear,
  newdata = test_data
)


# evaluation

confusionMatrix(
  svm_linear_pred,
  test_data$depressed,
  positive = "1"
)


# Radial SVM


set.seed(123)

svm_model <- svm(
  depressed ~
    sleep_cat +
    bed_time_weekday_hour +
    bed_time_weekend_hour +
    RIDAGEYR +
    RIAGENDR +
    DMDEDUC2 +
    marital3group +
    INDFMPIR +
    DMDBORN4,
  
  data = train_balanced,
  
  kernel = "radial",
  cost = 1,
  gamma = 0.1,
  
  scale = TRUE
)

# predictions
svm_pred <- predict(
  svm_model,
  newdata = test_data
)

# confusion matrix
confusionMatrix(
  svm_pred,
  test_data$depressed,
  positive = "1"
)



```


# polychoric PCA


```{r}
library(dplyr)
library(tidyr)
library(psych)

phq_items <- nhanes_all %>%
  filter(RIDAGEYR >= 20) %>%
  dplyr::select(
    DPQ010, DPQ020, DPQ030,
    DPQ040, DPQ050, DPQ060,
    DPQ070, DPQ080, DPQ090
  ) %>%
  mutate(across(
    everything(),
    ~ as.numeric(.)
  )) %>%
  mutate(across(
    everything(),
    ~ ifelse(. %in% c(7, 9, 77, 99), NA, .)
  )) %>%
  drop_na()

# check values should only be 0,1,2,3
lapply(phq_items, table)






```


# Paper style table


```{r}
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)

missing_table_paper <- missing_table %>%
  filter(Missing_Count > 0) %>%
  mutate(
    Missing_Rate = paste0(Missing_Rate, "%")
  ) %>%
  rename(
    `Variable` = Variable,
    `Missing Count` = Missing_Count,
    `Total N` = Total_N,
    `Missing Rate` = Missing_Rate
  )

kable(
  missing_table_paper,
  caption = "Table 1. Missingness Summary of Variables",
  align = c("l", "r", "r", "r"),
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    font_size = 10
  )

missing_table_paper <- missing_table %>%
  filter(Missing_Count > 0) %>%
  slice_max(Missing_Rate, n = 10) %>%
  mutate(
    Missing_Rate = paste0(Missing_Rate, "%")
  ) %>%
  rename(
    `Variable` = Variable,
    `Missing Count` = Missing_Count,
    `Total N` = Total_N,
    `Missing Rate` = Missing_Rate
  )

kable(
  missing_table_paper,
  caption = "Table 1. Missingness Summary of Variables",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    font_size = 10
  )



library(broom)

logit_table <- broom::tidy(log_model) %>%
  mutate(
    Odds_Ratio = exp(estimate),
    estimate = round(estimate, 3),
    Odds_Ratio = round(Odds_Ratio, 3),
    p.value = round(p.value, 4)
  ) %>%
  dplyr::select(
    term,
    estimate,
    Odds_Ratio,
    p.value
  ) %>%
  rename(
    Variable = term,
    Coefficient = estimate,
    `Odds Ratio` = Odds_Ratio,
    `P-value` = p.value
  )

kable(
  logit_table,
  caption = "Table 2. Logistic Regression Results",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  )


lasso_coef <- coef(lasso_cv, s = "lambda.min")

lasso_table <- data.frame(
  Variable = rownames(lasso_coef),
  Coefficient = as.numeric(lasso_coef)
) %>%
  filter(Coefficient != 0)

kable(
  lasso_table,
  caption = "Table 3. Non-Zero Coefficients Selected by LASSO",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  )





rf_importance <- importance(rf_model_bal)

rf_table <- data.frame(
  Variable = rownames(rf_importance),
  MeanDecreaseAccuracy = round(rf_importance[, "MeanDecreaseAccuracy"], 2),
  MeanDecreaseGini = round(rf_importance[, "MeanDecreaseGini"], 2)
) %>%
  arrange(desc(MeanDecreaseGini))

kable(
  rf_table,
  caption = "Table 5. Random Forest Variable Importance",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  )



model_results <- data.frame(
  Model = c(
    "Logistic Regression",
    "Decision Tree",
    "Balanced Random Forest",
    "Linear SVM",
    "Radial SVM"
  ),
  Accuracy = c(
    0.62,
    0.60,
    0.62,
    0.62,
    0.60
  ),
  Sensitivity = c(
    0.62,
    0.58,
    0.58,
    0.62,
    0.63
  ),
  Specificity = c(
    0.60,
    0.55,
    0.63,
    0.62,
    0.60
  ),
  Balanced_Accuracy = c(
    0.61,
    0.56,
    0.60,
    0.62,
    0.62
  )
)

kable(
  model_results,
  caption = "Table 6. Model Performance Comparison",
  booktabs = TRUE,
  digits = 3
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  )




svm_results <- data.frame(
  Model = c("Linear SVM", "Radial SVM"),
  Accuracy = c(0.622, 0.602),
  Sensitivity = c(0.620, 0.632),
  Specificity = c(0.622, 0.599),
  Balanced_Accuracy = c(0.621, 0.616)
)

kable(
  svm_results,
  caption = "Table 7. SVM Model Performance",
  booktabs = TRUE,
  digits = 3
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  )



```



MA 677 final assignment

```{r}


library(dplyr)
library(tidyr)
library(ggplot2)



dpq_vars <- grep("^DPQ", names(nhanes_all), value = TRUE)



nhanes_clean <- nhanes_all %>%
  mutate(
    across(all_of(dpq_vars), ~replace_na(., 0))
  )



nhanes_clean$depression_score <- rowSums(
  dplyr::select(nhanes_clean, all_of(dpq_vars)),
  na.rm = TRUE
)



nhanes_clean <- nhanes_clean %>%
  mutate(
    log_depression_score = log(depression_score + 1)
  )






p1 <- ggplot(nhanes_clean, aes(x = depression_score)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Depression Score",
    x = "Depression Score",
    y = "Count"
  ) +
  theme_minimal()



p2 <- ggplot(nhanes_clean, aes(x = log_depression_score)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Log Transformed Depression Score",
    x = "log(Depression Score + 1)",
    y = "Count"
  ) +
  theme_minimal()



p1
p2




library(MASS)


nhanes_clean <- nhanes_clean %>%
  mutate(
    sqrt_depression_score = sqrt(depression_score),
    log05_depression_score = log(depression_score + 0.5),
    log1_depression_score = log(depression_score + 1)
  )



plot_data <- nhanes_clean %>%
  dplyr::select(
    depression_score,
    log1_depression_score,
    log05_depression_score,
    sqrt_depression_score
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Transformation",
    values_to = "Value"
  )


ggplot(plot_data, aes(x = Value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~ Transformation, scales = "free") +
  labs(
    title = "Comparison of Depression Score Transformations",
    x = "Value",
    y = "Count"
  ) +
  theme_minimal()


```

different sample size

```{r}

library(dplyr)
library(ggplot2)

set.seed(123)

x <- nhanes_clean$depression_score
x <- as.numeric(x)
x <- x[!is.na(x)]

mu_hat <- mean(x)
sigma2_hat <- var(x)

sample_sizes <- c(30, 100)
B <- 100

results <- data.frame()

for (n in sample_sizes) {

  transformed_means <- replicate(B, {

    sample_x <- sample(x, size = n, replace = TRUE)

    xbar <- mean(sample_x)

    log(xbar + 1)
  })

  empirical_var <- var(transformed_means)

  delta_var <- (1 / (mu_hat + 1))^2 * (sigma2_hat / n)

  results <- rbind(
    results,
    data.frame(
      sample_size = n,
      empirical_variance = empirical_var,
      delta_variance = delta_var
    )
  )
}

results

```






